Efficient pedigree recording for fast population genetics simulation

### 2018-01-06 forwards_paper.tex
edited RAM section re: #44

The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  Pedigree tracking
dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
(Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, pedigree tracking reduces runtimes more for larger
population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).  We saw a similar trend with $4Nr$ for \simupop{}, but the effect of $N$
could not be explored because the simulations tended to run out of available memory.
  Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
partially obscures an important observation that the run times appear approximately linear in region size when tracking
pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on
average
each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.

\jda{Maybe we should quantify the factor by which RAM use increased; it's not
	easy to read off the figures}
\krt{I'm not sure it really matters.  It goes from a little to a lot.  And if we do relative, then we lose a lot of
    points again}
Our method uses more RAM (up to 2 orders of magnitude more) than simple forward simulations (see~\ref{ss:memuse}),
although in practice shorter simplification intervals may reduce RAM use
(but at a slight cost to the speed up achieved, see~\ref{ss:gcinterval}).
The benefits to our method are not limited to situations with selection:
in a more limited set of neutral simulations we found the same qualitative
behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:timing_nosel}).

>>>>

The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  Pedigree tracking
dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
(Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, pedigree tracking reduces runtimes more for larger
population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).  We saw a similar trend with $4Nr$ for \simupop{}, but the effect of $N$
could not be explored because the simulations tended to run out of available memory.
Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
partially obscures an important observation that the run times appear approximately linear in region size when tracking
pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative
behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:timing_nosel}).

In our test runs, our method used substantially more RAM than simple forward simulations (see~\ref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory. 
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see~\ref{ss:gcinterval}). 
In fact, our method could in some situations \emph{reduce} the amount of memory required, 
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.
